SunPy - Python for Solar Physics

### 2014-01-24 1-Introduction.tex
spellchecked now and changed the use of import for acquisition

Currently, most solar data analysis is performed with a large library 
of routines called SolarSoft, also referred to as SSW 
\cite{freeland1998}. SolarSoft is a set of integrated software 
libraries, databases, and system utilities which provide a common 
programming and data analysis environment for solar physics. 
SSW is structured by a package system which contains
specialised software for different instrument or tools, 
allowing the installation of just the needed packages.
It is 
overwhelmingly an environment which relies upon IDL (Interactive Data 
Language), a commercial and closed-source interactive data analysis 
environment sold by Exelis, a global aerospace, defense, information 
and services company. The SSW gen package provides the base analysis 
capabilities such as time series analysis, time conversions, spectral 
fitting, image display, and file I/O. It is composed of 89\% IDL code 
while the remainder is mostly perl (5.4\%) and csh scripts (3.8\%)
\footnote{{\bf This data was calculated using...}}. 
According to data generated using David A. Wheeler's 'SLOCCount'\footnote{
\url{http://www.dwheeler.com/sloccount/}}, 
SSW/gen is composed of 289,724 lines of code with a total approximate 
value of \$10.4M (based on the Basic COCOMO model \cite{_cocomo_2014}).
\footnote{{\bf This data was calculated using...} If we can add them to the
repository will improve the quality of the paper and we can always repeat the calculation in the future}. 
Individual 
missions and instruments provide additional optional packages to 
support and process their data. For example, the SDO package 
\cite{sdo} is 100\% IDL code and is 11k lines of code (\$360k) 
and the SOHO package \cite{soho} is 100k lines of code (\$3M) and is 
99.9\% IDL code. SSW is hosted and distributed by the Lockheed Martin 
Solar and Astrophysics Laboratory.
\footnote{Do we know whether they have some licence owning it? do we know the solarsoft licence at all?}
%The RHESSI package is 99.97\% IDL code and is 145k total lines of code (\$5M)
While SSW is open-source and freely available, the IDL core is not. The cost 
of IDL is prohibitive for many undergraduates as well as institutions 
with limited means.

One of SunPy's key aims is to provide a free and modern alternative 
to the SolarSoft (gen) library. This is made possible by the recent 
rise of the scientific Python community. Python is a general purpose 
high-level programming language. It is free to use, cross-platform 
and is both an accessible and powerful programming language. Python 
is one of the top ten most popular programming languages in the world 
according to the 2014 TIOBE Index 
\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}.
It therefore has a very large user community which extends faroutside 
of the solar physics or astronomy communities. It is very extensible 
with C, C++, Fortran or even IDL. Many books and on-line 
documentation is available. For example at the time writing a search
\footnote{We need to be careful with this! IDL means more things that
our beloved ``Interactive Data Language''. If we search for that
and programming then we get just 7 books! - from which 5 are just the same.
I don't know how we could make a more accurate search.
If we script it it would be amazing!  Also, I would just provide the order
of magnitude, and not a exact number, i.e. 190 and 1800 results} 
on Amazon for books on IDL programming generated $189$ results while 
a search for Python yielded $1\,788$ results. Python has a strong 
emphasis on readability a concept which is well-aligned with the 
scientific endeavor.
Additionally, Python is increasingly being taught in universities,
which means that research projects can begin quickly 
without the need for students to learn a new (and relatively niche) 
language. Finally, knowledge of Python is a skill which is useful and well 
compensated outside of the scientific community (average salary of 
\$110k in the Washington DC area according to indeed.com at the time 
of writing). Python continues to see increased use in the astronomy 
community \cite{greenfield2011} which has similar goals and 
requirements as the solar physics community. Finally, Python plays 
well with many technologies such as web servers \cite{dolgert2008}, 
SQL databases, interactive and shareable notebook-based computing 
\cite{perez2007}. For all of these reasons, Python is an ideal 
platform for a modern solar physics data analysis environment.

SunPy is built on top of the core scientific Python packages namely 
\texttt{NumPy}, \texttt{SciPy}, \texttt{matplotlib}. These core 
packages provide capabilities surpassing that provided by IDL for 
array manipulation and 2-d plotting \cite{greenfield2011}. Providing 
similar capabilities as SSW/gen in Python is made easier by 
leveraging the diverse Python community. For example, SunPy makes use 
of the \texttt{Pandas} package to provide high-performance, 
easy-to-use data structures and data analysis tools for time series 
data\cite{mckinney2012}. \texttt{Pandas} was originally developed for 
quantitative analysis of financial data and has seen grown into a 
generalized data analysis package. SunPy also relies on and is 
coordinating with the \texttt{Astropy} package 
\cite{theastropycollaboration2013}. It is still in active development 
but already provides many relevant capabilities such as full-featured
support for FITS files, absolute dates and time conversions, units 
and unit conversions, and model-fitting. Pandas (134k lines of code) 
and AstroPy (90k lines of code) allow SunPy to leverage a total value 
of \$7.6M worth of development which is close to the \$10.4M for 
SSW/gen assuming a rough equivalency between a line of Python code 
and a line of IDL code.

The design philosophy of SunPy is to provide a clean, simple to use, 
and well structured package that provides the \textit{core} tools for 
solar physics. The primary focus of SunPy's development early 
development is to provide specialised, linked, datatypes that allow 
the import, processing and visualisation of all types of solar data. 
These include \texttt{Map}, \texttt{LightCurve} and 
\texttt{Spectrum}, these cover spatial, timeseries and spectral data  
respectively.

>>>>

Currently, most solar data analysis is performed with a large library 
of routines called SolarSoft, also referred to as SSW 
\cite{freeland1998}. SolarSoft is a set of integrated software 
libraries, databases, and system utilities which provide a common 
programming and data analysis environment for solar physics. 
SSW is structured by a package system which contains
specialised software for different instrument or tools, 
allowing the installation of just the needed packages.
It is 
overwhelmingly an environment which relies upon IDL (Interactive Data 
Language), a commercial and closed-source interactive data analysis 
environment sold by Exelis, a global aerospace, defence, information 
and services company. The SSW gen package provides the base analysis 
capabilities such as time series analysis, time conversions, spectral 
fitting, image display, and file I/O. It is composed of 89\% IDL code 
while the remainder is mostly Perl (5.4\%) and csh scripts (3.8\%)
\footnote{{\bf This data was calculated using...}}. 
According to data generated using David A. Wheeler's 'SLOCCount'\footnote{
\url{http://www.dwheeler.com/sloccount/}}, 
SSW/gen is composed of 289,724 lines of code with a total approximate 
value of \$10.4M (based on the Basic COCOMO model \cite{_cocomo_2014}).
\footnote{{\bf This data was calculated using...} If we can add them to the
repository will improve the quality of the paper and we can always repeat the calculation in the future}. 
Individual 
missions and instruments provide additional optional packages to 
support and process their data. For example, the SDO package 
\cite{sdo} is 100\% IDL code and is 11k lines of code (\$360k) 
and the SOHO package \cite{soho} is 100k lines of code (\$3M) and is 
99.9\% IDL code. SSW is hosted and distributed by the Lockheed Martin 
Solar and Astrophysics Laboratory.
\footnote{Do we know whether they have some licence owning it? do we know the SolarSoft licence at all?}
%The RHESSI package is 99.97\% IDL code and is 145k total lines of code (\$5M)
While SSW is open-source and freely available, the IDL core is not. The cost 
of IDL is prohibitive for many undergraduates as well as institutions 
with limited means.

One of SunPy's key aims is to provide a free and modern alternative 
to the SolarSoft (gen) library. This is made possible by the recent 
rise of the scientific Python community. Python is a general purpose 
high-level programming language. It is free to use, cross-platform 
and is both an accessible and powerful programming language. Python 
is one of the top ten most popular programming languages in the world 
according to the 2014 TIOBE Index 
\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}.
It therefore has a very large user community which extends far outside 
of the solar physics or astronomy communities. It is very extensible 
with C, C++, FORTRAN or even IDL. Many books and on-line 
documentation is available. For example at the time writing a search
\footnote{We need to be careful with this! IDL means more things that
our beloved ``Interactive Data Language''. If we search for that
and programming then we get just 7 books! - from which 5 are just the same.
I don't know how we could make a more accurate search.
If we script it it would be amazing!  Also, I would just provide the order
of magnitude, and not a exact number, i.e. 190 and 1800 results} 
on Amazon for books on IDL programming generated $189$ results while 
a search for Python yielded $1\,788$ results. Python has a strong 
emphasis on readability a concept which is well-aligned with the 
scientific endeavour.
Additionally, Python is increasingly being taught in universities,
which means that research projects can begin quickly 
without the need for students to learn a new (and relatively niche) 
language. Finally, knowledge of Python is a skill which is useful and well 
compensated outside of the scientific community (average salary of 
\$110k in the Washington DC area according to indeed.com at the time 
of writing). Python continues to see increased use in the astronomy 
community \cite{greenfield2011} which has similar goals and 
requirements as the solar physics community. Finally, Python plays 
well with many technologies such as web servers \cite{dolgert2008}, 
SQL databases, interactive and shareable notebook-based computing 
\cite{perez2007}. For all of these reasons, Python is an ideal 
platform for a modern solar physics data analysis environment.

SunPy is built on top of the core scientific Python packages namely 
\texttt{NumPy}, \texttt{SciPy}, \texttt{matplotlib}. These core 
packages provide capabilities surpassing that provided by IDL for 
array manipulation and 2-d plotting \cite{greenfield2011}. Providing 
similar capabilities as SSW/gen in Python is made easier by 
leveraging the diverse Python community. For example, SunPy makes use 
of the \texttt{Pandas} package to provide high-performance, 
easy-to-use data structures and data analysis tools for time series 
data\cite{mckinney2012}. \texttt{Pandas} was originally developed for 
quantitative analysis of financial data and has seen grown into a 
generalised data analysis package. SunPy also relies on and is 
coordinating with the \texttt{Astropy} package 
\cite{theastropycollaboration2013}. It is still in active development 
but already provides many relevant capabilities such as full-featured
support for FITS files, absolute dates and time conversions, units 
and unit conversions, and model-fitting. Pandas (134k lines of code) 
and Astropy (90k lines of code) allow SunPy to leverage a total value 
of \$7.6M worth of development which is close to the \$10.4M for 
SSW/gen assuming a rough equivalency between a line of Python code 
and a line of IDL code.

The design philosophy of SunPy is to provide a clean, simple to use, 
and well structured package that provides the \textit{core} tools for 
solar physics. The primary focus of SunPy's development early 
development is to provide specialised, linked, datatypes that allow 
the acquisition, processing and visualisation of all types of solar data. 
These include \texttt{Map}, \texttt{LightCurve} and 
\texttt{Spectrum}, these cover spatial, timeseries and spectral data  
respectively.
